---
title: "Data Visualization with ggplot2"
author: "Jenn Schilling"
date: "May 25, 2022"
output:
 html_document:
  theme: "flatly" 
  toc: TRUE
  toc_float: TRUE
---


This file contains contains code for a ResBaz Tucson Workshop: Data Visualization with ggplot2, presented by Jenn Schilling on May 25, 2022. 

**Materials developed by Jenn Schilling.**  


# Setup 

```{r setup, message = FALSE, warning = FALSE}

# Set default chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE) 

# Set the default size of figures
knitr::opts_chunk$set(fig.width = 8, fig.height = 5)  

# Load libraries
library(here)
library(palmerpenguins)
library(tidyverse)
library(scales)

```


# View the data

```{r read-data}

# Data from Palmer Penguins
View(penguins)

```


# Data Visualization

## Scatterplot

```{r data-viz-01-scatter}

# Is there a relationship between bill length and bill depth?

# Step 1: Specify the data
ggplot(data = penguins)

# Step 2: Specify the aesthetic mapping
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm))

# Step 3: Specify the geometric object
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point()

# Step 4: Add to aesthetic mapping

# Categorical color
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     color = species)) +
  geom_point()

# Shape
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     shape = species)) +
  geom_point()

# Sequential color
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     color = flipper_length_mm)) +
  geom_point()

# Size
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     size = flipper_length_mm)) +
  geom_point()


```

## Line Graph

```{r data-viz-02-line}

# How does the number of penguins change over time?

# This time, we need to do a little processing
penguins_time <- penguins %>%
  count(species, year)

# Set up the data, aesthetic mapping, and geometric objects
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n)) +
  geom_line()

# Something doesn't look right - there is only one line

# Need to add the group argument in the aes() mapping to tell R to create one line per group (institution)
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species)) +
  geom_line()

# Add Color
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line()

# Add Labels
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species,
                     label = species)) +
  geom_line() +
  geom_text()

# Add labels to the end of the line
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line() +
  geom_text(data = penguins_time %>% 
              filter(year == 2009),
            mapping = aes(x = year,
                          y = n,
                          color = species,
                          label = species))

# Adjust label position
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line() +
  geom_text(data = penguins_time %>% 
              filter(year == 2009),
            mapping = aes(x = year,
                          y = n,
                          color = species,
                          label = species),
            hjust = 0) + # Step 1: Align Right
  coord_cartesian(clip = "off") + # Step 2: Turn off clipping
  guides(color = "none") + # Step 3: Remove legend since we no longer need it
  theme(plot.margin = margin(10, 30, 10, 10)) # Step 4: Adjust plot margin since Step 3 caused labels to cut off again



```

## Bar Graph

```{r data-viz-03-bar}

# How many penguins of each species?

# Set up the data, aesthetic mapping, and geometric objects
ggplot(data = penguins,
       mapping = aes(x = species)) +
  geom_bar()

# geom_bar() computes counts for us - we just need to provide the x-axis category

# Can we sort them?
# Yes, by using the fct_infreq() function from the forcats package in tidyverse
# fct_infreq() reorders the levels by their frequencies
ggplot(data = penguins,
       mapping = aes(x = fct_infreq(species))) +
  geom_bar()

# What if we want to make a bar plot with something other than counts?

# What is the average bill length for each species?

# Create a new dataset with average bill length for each species
penguins_avg <- penguins %>%
  group_by(species) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))

# Plot average bill length by species
ggplot(data = penguins_avg,
       mapping = aes(x = species,
                     y = avg_bill_length)) +
  geom_col() # geom_col() plots a bar with the data specified in the y argument of aes()

# Now plot the aggregated data and arrange the bars in order 
# This time, use the reorder() function to arrange the species because we have a numeric value we want to use for the ordering
ggplot(data = penguins_avg,
       mapping = aes(x = reorder(species, -avg_bill_length),
                     y = avg_bill_length)) +
  geom_col()

# How do we add color?
ggplot(data = penguins_avg,
       mapping = aes(x = reorder(species, -avg_bill_length),
                     y = avg_bill_length,
                     fill = species)) +
  geom_col()

```

## Pie Chart

```{r data-viz-04-pie}

# What proportion of penguins are male vs. female?

# Pick on species - "Gentoo"

# Plot the data, aesthetics mapping, and geometric object
ggplot(data = penguins %>%
         filter(species == "Gentoo"),
       mapping = aes(x = species,
                     fill = sex)) +
  geom_bar()

# Change the coordinate system to make a pie chart
ggplot(data = penguins %>%
         filter(species == "Gentoo"),
       mapping = aes(x = species,
                     fill = sex)) +
  geom_bar() +
  coord_polar(theta = "y")

# What if we want to use percentages?

# Make a new data table with the percentages
penguin_perc <- penguins %>%
  filter(species == "Gentoo") %>%
  count(sex) %>%
  mutate(perc = n / sum(n))

# Plot the percentages
ggplot(data = penguin_perc,
       mapping = aes(x = "",
                     y = perc,
                     fill = sex)) +
  geom_col() +
  coord_polar(theta = "y",
              start = 0) 

# Add labels
ggplot(data = penguin_perc,
       mapping = aes(x = "",
                     y = perc,
                     fill = sex,
                     label = percent(perc))) +
  geom_col() +
  geom_text(position = position_fill(vjust = 0.5)) +
  coord_polar(theta = "y",
              start = 0) 


```

## Faceting


```{r data-viz-05-facets}

# How many penguins of each species are on different islands?

# Plot the data, aesthetic mapping, and geometric object
ggplot(data = penguins,
       mapping = aes(y = species)) +
  geom_bar()

# We could add color to see the island
ggplot(data = penguins,
       mapping = aes(y = species,
                     fill = island)) +
  geom_bar()

# We could separate out the stacked bars
ggplot(data = penguins,
       mapping = aes(y = species,
                     fill = island)) +
  geom_bar(position = "dodge")

# Or we could make small charts for each island
ggplot(data = penguins,
       mapping = aes(y = species,
                     fill = island)) +
  geom_bar()+
  facet_wrap(~ island) # Make a small plot for each island

# Make some formatting changes for the facets
ggplot(data = penguins,
       mapping = aes(y = species,
                     fill = island)) +
  geom_bar()+
  facet_wrap(~ island,
             ncol = 1, # Put all the plots in one column
             scales = "free_y") # Let the y-axis be "free" so only the relevant institutions to that locale show


```

## Facets with Bars and Lines

```{r data-viz-06-bar}

# How many penguins are there over time of each species (and on each island)?

# Plot - this creates a stacked bar graph
ggplot(data = penguins,
       mapping = aes(x = year,
                     fill = species)) +
  geom_bar()

# Plot - this creates a side-by-side bar graph
ggplot(data = penguins,
       mapping = aes(x = year,
                     fill = species)) +
  geom_bar(position = "dodge")

# Make a small plot for each species
ggplot(data = penguins,
       mapping = aes(x = year,
                     fill = species)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ species)

# Remove the legend since we no longer need it
ggplot(data = penguins,
       mapping = aes(x = year,
                     fill = species)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ species) +
  guides(fill = "none")

# Change to faceting by island and add the legend back in
ggplot(data = penguins,
       mapping = aes(x = year,
                     fill = species)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ island)

# Since we're showing data over time, maybe a line graph would be better

# First we need to create a new dataset that is aggregated
penguins_time_2 <- penguins %>%
  count(species, island, year)

# Can we just replace geom_col() with geom_line()?
ggplot(data = penguins_time_2,
       mapping = aes(x = year,
                     y = n,
                     fill = species)) +
  geom_line() +
  facet_wrap(~ island) 

# Where did the color go? Also, notice the change in the y-axis - it no longer starts at 0

# Replace "fill" with "color" to get the line color back
ggplot(data = penguins_time_2,
       mapping = aes(x = year,
                     y = n,
                     color = species)) +
  geom_line() +
  facet_wrap(~ island) 

# Bars, histograms, and boxplots use fill; lines and points use color

```


## Formatting and Saving

```{r data-viz-07-format}

# So far, we’ve gone over a few different plot types, how to add color, and faceting. But how do you edit the colors and theme of the plot to get something that looks really nice? We’ve seen some pretty strange and unhelpful axis labels. How do we fix that?

# Let's return to: How does the number of penguins change over time?

# We've already made some adjustments to this plot
# We removed the legend and added labels to the end of the line (explain why this helps)

p <- ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line() +
  geom_text(data = penguins_time %>% 
              filter(year == 2009),
            mapping = aes(x = year,
                          y = n,
                          color = species,
                          label = species),
            hjust = 0 #, nudge_x = 0.015 # Add this in at the end to move labels just slightly away from points
            ) + 
  coord_cartesian(clip = "off") + 
  guides(color = "none") + 
  theme(plot.margin = margin(10, 30, 10, 10)) 

p

# Things we want to fix:
# 1. x-axis labels should not have decimal place
# 2. Axis titles are not very helpful
# 3. There's not plot title or data source noted
# 4. Use colors other than the R defaults for the lines
# 5. Remove the grey background from the plot


# 1. x-axis labels should not have decimal place
# Adjust the scale for the x-axis
p_1 <- p +
  scale_x_continuous(labels = number_format(big.mark = "", accuracy = 1))

p_1

# 3. Axis titles are not very helpful
# 4. There's not plot title or data source noted
# Add on the labs() function
# Mention color and fill arguments
# Talk about better plot title
p_2 <- p_1 +
  labs(title = "Number of penguins observed each year",
       x = "Year",
       y = "", 
       caption = "Data Source: Palmer Penguins R Package")

p_2

# 5. Use colors other than the R defaults for the lines
# This is another application of a scale function
p_3 <- p_2 +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) 

p_3

# We can also use other pre-built color schemes 
p_2 +
  scale_color_brewer(palette = "Set1") 

# Or we can use HEX codes
 p_2 +
  scale_color_manual(values = c( "#53868b", "#f18805", "#3f5a00")) 

# 6. Remove the grey background from the plot
p_4 <- p_3 +
  theme_minimal() 

p_4

# Notice our plot margin extension was removed
# So, add that back in
p_5 <- p_4 +
  theme(plot.margin = margin(10, 30, 10, 10)) 

p_5

# Final touch-ups
# Add a point for each year
p_6 <- p_5 + 
  geom_point()

p_6

# Save the plot
ggsave(filename = here("plots", "line_plot.png"),
       plot = p_6,
       device = "png",
       width = 8,
       height = 5,
       type = "cairo") # smoother lines

```

## Fixing Potential Errors

```{r data-viz-08-erro, eval = FALSE}

# Why doesn't the code in this chunk create what we expect?

## 1 -  Color ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     color = "blue")) +
  geom_point()


# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point(color = "blue")

## 2 - Plus sign ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm))
 + geom_point() # <- + at the beginning


ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) # <- missing +
  geom_point()
  
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) + + # <- double +
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) + # Need to have a single + at the end of the line
  geom_point()


## 3 - Missing object ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth,
                     y = bill_length_mm)) +
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point()

## 4 - Parentheses ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)+ # <- missing parentheses
  geom_point()
  
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm))) + # <- extra parentheses 
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point()

```


## Data Visualization - Your Turn

Using the code we've already written, try making a new version of a plot. Use a different geometric object (i.e. replace `geom_line()` with `geom_bar()`) or change the aesthetics mappings (i.e. add a color or fill argument to the `aes()` function or switch out the variables used in the `aes()` function to visualize different values). 

Learn more at the package website: https://ggplot2.tidyverse.org/ 


```{r visualize-data-your-turn}




```


