---
title: "Data Visualization with ggplot2"
author: "Jenn Schilling"
date: "May 25, 2022"
output:
 html_document:
  theme: "flatly" 
  toc: TRUE
  toc_float: TRUE
---


This file contains contains code for a ResBaz Tucson Workshop: Data Visualization with ggplot2, presented by Jenn Schilling on May 25, 2022. 

**Materials developed by Jenn Schilling.**  

# Setup

```{r setup}

# Set default chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE) 

# Set the default size of figures
knitr::opts_chunk$set(fig.width = 8, fig.height = 5)  

# Install packages - if needed
# install.packages(c("here", "palmerpenguins", "tidyverse", "scales"))

# Load libraries
library(here)
library(palmerpenguins)
library(tidyverse)
library(scales)

```


# View the data

```{r read-data}

# Data from Palmer Penguins
View(penguins)

```


# Data Visualization

## Scatterplot

```{r data-viz-01-scatter}

# Is there a relationship between bill length and bill depth?


# Step 1: Specify the data
ggplot(data = penguins)

# Step 2: Specify the aesthetic mapping
ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm,
                     y = bill_depth_mm))

# Step 3: Specify the geometric object
ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm,
                     y = bill_depth_mm)) +
  geom_point()

# Step 4: Add to aesthetic mapping
ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm,
                     y = bill_depth_mm,
                     color = species)) +
  geom_point()






```

## Line Graph

```{r data-viz-02-line}

# How does the number of penguins change over time?

# Get the number of penguins of each species and year in a table
penguins_time <- penguins %>%
  count(species, year)

# Create our plot
ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line() +
  geom_text(data = penguins_time %>% filter(year == 2009),
            mapping = aes(label = species),
            hjust = 0) + # Right aligned the labels
  coord_cartesian(clip = "off") + # Turned off clipping so we can see the labels
  guides(color = "none") + # Remove color legend
  theme(plot.margin = margin(10, 30, 10, 10)) + # Adding padding to the margin for the labels
  labs(x = "Year",
       y = "Number of Penguins",
       caption = "Data from the Palmer Penguins R Package",
       title = "Number of Penguins by Year",
       subtitle = "One line per penguin speices")



```

## Bar Graph

```{r data-viz-03-bar}

# How many penguins of each species?
ggplot(data = penguins,
       mapping = aes(x = species)) +
  geom_bar()

# What if we wanted to order the species by count?
ggplot(data = penguins,
       mapping = aes(x = fct_rev(fct_infreq(species)))) +
  geom_bar()

# What is the average bill length for each species?

# Computed the average bill length
penguins_avg <- penguins %>%
  group_by(species) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))

penguin_colors <- c("#53868b", "#f18805", "#3f5a00")

# Plot in a bar graph
ggplot(data = penguins_avg,
       mapping = aes(x = reorder(species, -avg_bill_length),
                     y = avg_bill_length,
                     fill = species)) +
  geom_col() + # plot values directly as bars
  scale_fill_manual(values = penguin_colors)

```

## Pie Chart

```{r data-viz-04-pie}

# What proportion of penguins are male vs. female?


```

## Faceting

```{r data-viz-05-facets}

# How many penguins of each species are on different islands?


```

# Facets with Bars and Lines

```{r data-viz-06-bar}

# How many penguins are there over time of each species (and on each island)?


```

## Formatting and Saving

```{r data-viz-07-format}

# So far, we’ve gone over a few different plot types, how to add color, and faceting. But how do you edit the colors and theme of the plot to get something that looks really nice? We’ve seen some pretty strange and unhelpful axis labels. How do we fix that?

# Let's return to: How does the number of penguins change over time?
p <- ggplot(data = penguins_time,
       mapping = aes(x = year,
                     y = n,
                     group = species,
                     color = species)) +
  geom_line() +
  geom_text(data = penguins_time %>% filter(year == 2009),
            mapping = aes(label = species),
            hjust = 0) + # Right aligned the labels
  coord_cartesian(clip = "off") + # Turned off clipping so we can see the labels
  guides(color = "none") + # Remove color legend
  theme(plot.margin = margin(10, 30, 10, 10)) + # Adding padding to the margin for the labels
  labs(x = "Year",
       y = "Number of Penguins",
       caption = "Data from the Palmer Penguins R Package",
       title = "Number of Penguins by Year",
       subtitle = "One line per penguin speices")

p

# We've already made some adjustments to this plot
# We removed the legend and added labels to the end of the line 
# We've also already added titles

# Put the year axis labels in whole numbers 
p_1 <- p +
  scale_x_continuous(breaks = c(2007, 2008, 2009))

p_1

# Change the colors
p_2 <- p_1 +
  scale_color_manual(values = c("darkorange", "purple", "cyan4"))

p_2

# Change the theme
p_3 <- p_2 +
  theme_minimal()

p_3

# Extend the right margin again
p_4 <- p_3 +
  theme(plot.margin = margin(10, 30, 10, 10)) 

p_4

# Save plot 
ggsave(filename = "line_plot.png",
       plot = p_4,
       device = "png",
       type = "cairo",
       width = 8,
       height = 5)

```

## Fixing Potential Errors

```{r data-viz-08-error}

# Why doesn't the code in this chunk create what we expect?

## 1 -  Color ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm,
                     color = "blue")) +
  geom_point()


# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point(color = "blue")

## 2 - Plus sign ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm))
 + geom_point() # <- + at the beginning


ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) # <- missing +
  geom_point()
  
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) + + # <- double +
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) + # Need to have a single + at the end of the line
  geom_point()


## 3 - Missing object ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth,
                     y = bill_length_mm)) +
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point()

## 4 - Parentheses ##

# Not quite right
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)+ # <- missing parentheses
  geom_point()
  
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm))) + # <- extra parentheses 
  geom_point()

# Correct
ggplot(data = penguins,
       mapping = aes(x = bill_depth_mm,
                     y = bill_length_mm)) +
  geom_point()

```


## Data Visualization - Your Turn

Using the code we've already written, try making a new version of a plot. Use a different geometric object (i.e. replace `geom_line()` with `geom_bar()`) or change the aesthetics mappings (i.e. add a color or fill argument to the `aes()` function or switch out the variables used in the `aes()` function to visualize different values). 

Learn more at the package website: https://ggplot2.tidyverse.org/ 

```{r visualize-data-your-turn}




```


